


\textit{A special thank to professor Jérôme Adda, who inspired me everyday in class, and taught me something that is found in this thesis, and goes way beyond it. 
}
\\

\textit{I thank professor David Rossell, who first sparked in me the curiosity for machine learning and helped crafting the idea for this thesis, and Lorenzo M, who gave me precious suggestions on these matters and much more. }

\textit{I am also deeply grateful to professors Massimo Marinacci, Maristella Botticini and Chiara Fumagalli: I am honoured to say that you gave me some advice I will always keep with me. }\\

\textit{A heartfelt thank to my friends Lucrezia, Bianca, Ilaria, Lorenzo G, Francesco S, Francesco D, Luigi, Federico, Gabriele N, Alessandro, Zoe, Michelle and Gabriele P. You made these three years better than I could have ever imagined. Thank you for keeping me going every time I don't see how or don't know where to go; I found in you another home.}

\textit{
Thank you also to my best friends Chiara, Roberta, Celeste and Anna, who will always remain such, no matter where life brings us.}

\textit{My last thanks to Elisa and Paola, without whom I sincerely think I would be completely lost. You are my role models and I feel incredibly grateful to have you by my side and be by yours for all the life to come.}\\


\textit{This thesis is dedicated to my parents Giovanna and Fabio, the best team I know. You give me an incredible strength to navigate life, while reminding me not to take it too seriously.}


